Type-Safe Modular Parsing

### 2016-12-21 paper/paper.tex
Update the tex template. Minor comments on Overview.

%\documentclass[preprint]{sigplanconf}
%\documentclass[preprint]{llncs}
\documentclass[a4paper,UKenglish]{lipics}

@@

\author[1]{John Q. Open}
\author[2]{Joan R. Access}
\affil[1]{Dummy University Computing Laboratory\\
  Address, Country\\
  \texttt{open@dummyuni.org}}
\affil[2]{Department of Informatics, Dummy College\\
  Address, Country\\
  \texttt{access@dummycollege.org}}
\authorrunning{J.\,Q. Open and J.\,R. Access}
% mandatory. First: Use abbreviated first/middle names. Second (only in severe
% cases): Use first author plus 'et. al.'

\Copyright{John Q. Open and Joan R. Access}
% mandatory, please use full first names. LIPIcs license is "CC-BY";
% http://creativecommons.org/licenses/by/3.0/

\subjclass{Dummy classification -- please refer to
  \url{http://www.acm.org/about/class/ccs98-html}}
% mandatory: Please choose ACM 1998 classifications from
% http://www.acm.org/about/class/ccs98-html . E.g., cite as "F.1.1 Models of
% Computation".
\keywords{Dummy keyword -- please provide 1--5 keywords}% mandatory: Please provide 1-5 keywords
% Author macros::end %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\serieslogo{}%please provide filename (without suffix)
\volumeinfo%(easychair interface)
  {Billy Editor and Bill Editors}% editors
  {2}% number of editors: 1, 2, ....
  {Conference title on which this volume is based on}% event
  {1}% volume
  {1}% issue
  {1}% starting page number
\EventShortName{}
\DOI{10.4230/LIPIcs.xxx.yyy.p}% to be completed by the volume editor
% Editor-only macros::end %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

  This paper investigates solutions for the problem of \emph{modular
    parsing}. We focus on \emph{semantic} modularity and not just
  \emph{syntactic} modularity. That is, the solutions should not only
  allow complete parsers to be built out of modular parsing
  components, but also enable the parsing components to be
  \emph{modularly type-checked} and \emph{separately compiled}. We
  present \name: a parsing combinator library that enables modular
  parsing. Interestingly, the modularity requirements for modular
  parsing rule out several existing parsing combinator approaches,
  which rely on some non-modular techniques. We show that Packrat
  parsing techniques, provide solutions for such modularity problems,
  and enable reasonable performance in a modular setting.  \name
  builts on top of a Packrat parsing library, but adds new parsing
  combinators to enable modular parsing. The new parsing combinators
  employ \emph{delegation-based} techniques and \emph{Object Algebras}
  to support extensibility. To evaluate \name we conduct a case study
  based on the ``Types and Programming Languages'' interpreters. The
  case study shows that \name is effective at reusing parsing code
  from existing interpreters, and the total parsing code is 60\%
  shorter than the non-modular parsing code.

>>>>

%\documentclass[preprint]{sigplanconf}
%\documentclass[preprint]{llncs}
\documentclass[a4paper,UKenglish]{lipics-v2016}

@@

%% Please provide for each author the \author and \affil macro, even when authors have the same affiliation, i.e. for each author there needs to be the  \author and \affil macros
\author[1]{John Q. Open}
\author[2]{Joan R. Access}
\affil[1]{Dummy University Computing Laboratory, Address/City, Country\\
  \texttt{open@dummyuniversity.org}}
\affil[2]{Department of Informatics, Dummy College, Address/City, Country\\
  \texttt{access@dummycollege.org}}
\authorrunning{J.\,Q. Open and J.\,R. Access} %mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et. al.'

\Copyright{John Q. Open and Joan R. Access}%mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/

\subjclass{Dummy classification -- please refer to \url{http://www.acm.org/about/class/ccs98-html}}% mandatory: Please choose ACM 1998 classifications from http://www.acm.org/about/class/ccs98-html . E.g., cite as "F.1.1 Models of Computation".
\keywords{Dummy keyword -- please provide 1--5 keywords}% mandatory: Please provide 1-5 keywords
% Author macros::end %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\EventEditors{John Q. Open and Joan R. Acces}
\EventNoEds{2}
\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
\EventShortTitle{CVIT 2016}
\EventAcronym{CVIT}
\EventYear{2016}
\EventDate{December 24--27, 2016}
\EventLocation{Little Whinging, United Kingdom}
\EventLogo{}
\SeriesVolume{42}
\ArticleNo{23}
% Editor-only macros::end %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

  This paper investigates solutions for the problem of \emph{modular
    parsing}. We focus on \emph{semantic} modularity and not just
  \emph{syntactic} modularity. That is, the solutions should not only
  allow complete parsers to be built out of modular parsing
  components, but also enable the parsing components to be
  \emph{modularly type-checked} and \emph{separately compiled}. We
  present \name: a parsing combinator library that enables modular
  parsing. Interestingly, the modularity requirements for modular
  parsing rule out several existing parsing combinator approaches,
  which rely on some non-modular techniques. We show that Packrat
  parsing techniques, provide solutions for such modularity problems,
  and enable reasonable performance in a modular setting.  Thus \name
  is built on top of a Packrat parsing library, but adds new parsing
  combinators to enable modular parsing. The new parsing combinators
  employ \emph{delegation-based} techniques and \emph{Object Algebras}
  to support extensibility. To evaluate \name we conduct a case study
  based on the ``Types and Programming Languages'' interpreters. The
  case study shows that \name is effective at reusing parsing code
  from existing interpreters, and the total parsing code is 60\%
  shorter than the non-modular parsing code.

### 2016-12-21 paper/sections/Overview.tex
Update the tex template. Minor comments on Overview.

This section gives an overview of our library \name, from a user's perspective. Basically, \name consists of four parts: underlying parsing technique, delegation mechanism encoded by open recursion, Object Algebras, and glue code of new combinators and utility functions. We start from section \ref{subsec:overview-parsing}, which discusses the choice of parsing technique and how it affects modularity of parsers. Section \ref{subsec:overview-problem} demonstrates the goal of extending parsers together with ASTs in a semantic modular way, with both separate compilation and type-safe code reuse. Then we will see traditional parser combinators fail to achieve it because of hard-coded recursive calls. In section \ref{subsec:overview-delegation}, we show how delegation can solve this problem and allow us to build extensible parsers. Finally, section \ref{subsec:overview-oa} gives examples of using Object Algebras for more extensibility, including extension of operations and parsing multiple sorts of syntax.

@@

\paragraph{Packrat Parsing}
Fortunately more advanced parsing techniques such as Packrat
parsing~\cite{Ford2002}, address the limitations of simple parser combinators
such as Parsec. Packrat parsers use
memoization to record the result of applying each parser at each
position of the input, so that repeated computation is eliminated, and
they support left-recursive grammars directly~\cite{warth2008}. All of these properties are very
suitable for modularity, thus we use Packrat parsers as the underlying
parsing technique. In section \ref{sec:packratparsers}, we will give
an overview of the Packrat parsing library implemented in Scala,
which we use.

@@

Typical parsers defined parser combinators follow the grammar
structure, and build a corresponding AST. If the grammar is changed to
add a new alternative, the corresponding parser has to be changed as
well. Thus the traditional style of developing parsers is not
extensible, and it is at odds with object-oriented programming ASTs,
which support the easy addition of new alternative classes. We
illustrate this problem with a simple expression language and extension.

@@

It is straightforward to model ASTs by inheritance and write corresponding parsers for each case. \huang{This is the first time to demonstrate code in this paper. Readers may not know what those combinators/types are. Refer to the library docs here? move the 'scala parsing library' section to the beginning as 'background'?}

@@

As the combination of \inlinecode{pVar} and \inlinecode{pApp}, \inlinecode{pExpr} can parse expressions correctly.

\paragraph{Attempt to Extend the Parser} Now consider we are extending the abstract syntax as well as the parsers. We add lambda abstractions as a new case of expressions. The language then becomes the famous \textit{untyped lambda calculus}.

@@

The ASTs can be extended trivially as below.

@@

\lstinputlisting[linerange=26-32]{../Scala/Parser/src/PaperCode/Overview/Overview.scala}% APPLY:linerange=OVERVIEW_SIMPLE_EXT

@@

\paragraph{Problem: Hard-coded Recursive Calls} The critical problem of extending parsers directly is that the recursive calls in old parsers are hard-coded. Specifically, \inlinecode{pApp} makes two recursive calls to parse two consecutive expressions as an application. Before the extension, we implemented it by using \inlinecode{pExpr} for those recursive calls. It worked fine at that time because \inlinecode{pExpr} covers all cases of the old grammar. However, after the extension of lambda abstractions, the proper function for such recursive calls should be \inlinecode{pExtExpr} instead of \inlinecode{pExpr}, because the new function \inlinecode{pExtExpr} covers all of the three cases in the new grammar, including the lambda case.

That is the reason why the extended parser fails to parse some valid expressions such as \inlinecode{(\\x.x)} \inlinecode{(\\y.y)}. To resolve this issue, we need to let \inlinecode{pApp} be aware of the change of grammar. We have some direct solutions:

\begin{itemize}
    \item Keep the old code, rewrite problematic functions like \inlinecode{pApp}. \\Separate compilation is obtained but code reuse is lost, because every function may contain such recursive calls.
    \item Modify the old code, correct the recursive calls by proper functions. \\We can avoid code duplication thus get code reuse, but separate compilation is lost because of modification on the old code.
    \item ??? \huang{third one? use dynamic languages?}
\end{itemize}

Similar to the trade-off in the Expression Problem, neither of them satisfies us. The challenge of building semantically modular parsers requires fundamental change of the parser structure. We will introduce our solution in the next sections.

@@

Hard-coded recursive calls prevent our parsers from extensibility.
This problem can be solved using delegation, which can be encoded with
\textit{open recursion}. The idea is simple: instead of making direct
recursive calls, an additional argument, that abstracts over the
recursive call, is passed to the parsers. This technique enables the
developement of parsers that support OO-style extensibility of ASTs.

@@

\lstinputlisting[linerange=53-63]{../Scala/Parser/src/PaperCode/Overview/Overview.scala}% APPLY:linerange=OVERVIEW_OPEN_BASE

According to the definition of type synonym \inlinecode{Open[T]},
every parse function takes a parser of \inlinecode{Expr} as a
parameter, namely \inlinecode{self}. This parameter \inlinecode{self}
represents \emph{the parser} of expressions, which is dynamiclly
explained by the argument. Using an abstract parameter provides
extensibility to parsers, because the recursive calls are not
hard-coded here --- they are dynamic and can be changed corresponding
to the grammar!

@@

\paragraph{Client Code} Since the type of parsing functions are
changed, we can not apply them directly. A \inlinecode{fix} function, which is the
the standard \textit{fixpoint combinator}, is used to `close' the open
recursion. After that, it can be used as an ordinary parser.
The client code below demonstrates the usage of new parsers.

@@

\lstinputlisting[linerange=68-72]{../Scala/Parser/src/PaperCode/Overview/Overview.scala}% APPLY:linerange=OVERVIEW_OPEN_USE

@@

\paragraph{Extensibility} As same as what we did in previous section, we extend the grammar by adding lambda abstractions as a new case. Similarly, the new parser extends from the old one by adding \inlinecode{pLam} on it. Notice that the old code is not touched, hence separate compilation is obtained.

\lstinputlisting[linerange=77-83]{../Scala/Parser/src/PaperCode/Overview/Overview.scala}% APPLY:linerange=OVERVIEW_OPEN_EXT

@@

\lstinputlisting[linerange=89-89]{../Scala/Parser/src/PaperCode/Overview/Overview.scala}% APPLY:linerange=OVERVIEW_OPEN_EXT_USE

As we demonstrated, delegation encoded by open recursion is the key technique to obtain semantic modularity. It enables type-safe code reuse and separate compilation in our parsers. We will discuss this topic further in section \ref{sec:openandparsing}.

>>>>

This section gives an overview of our library \name, from a user's perspective. Basically, \name consists of four parts: underlying parsing technique, delegation mechanism encoded by open recursion, Object Algebras, and glue code of new combinators and utility functions. We start from Section \ref{subsec:overview-parsing}, which discusses the choice of parsing technique and how it affects modularity of parsers. Section \ref{subsec:overview-problem} demonstrates the goal of extending parsers together with ASTs in a semantic modular way, with both separate compilation and type-safe code reuse. Then we will see traditional parser combinators fail to achieve it because of hard-coded recursive calls. In Section \ref{subsec:overview-delegation}, we show how delegation can solve this problem and allow us to build extensible parsers. Finally, Section \ref{subsec:overview-oa} gives examples of using Object Algebras for more extensibility, including extension of operations and parsing multiple sorts of syntax.

@@

\paragraph{Packrat Parsing}
Fortunately more advanced parsing techniques such as Packrat
parsing~\cite{Ford2002}, address the limitations of simple parser combinators
such as Parsec. Packrat parsers use
memoization to record the result of applying each parser at each
position of the input, so that repeated computation is eliminated, and
they support left-recursive grammars directly~\cite{warth2008}. All of these properties are very
suitable for modularity, thus we use Packrat parsers as the underlying
parsing technique. In Section \ref{sec:packratparsers}, we will give
an overview of the Packrat parsing library implemented in Scala,
which we use.

@@

Typical parsers defined by parser combinators follow the grammar
structure, they parse an input to build a corresponding AST. If the grammar is extended
with a new alternative, the corresponding parser has to be changed as
well. Thus the traditional style of developing parsers is not
extensible, and it is at odds with object-oriented programming ASTs,
which support the easy addition of new alternative classes. We
illustrate this problem with a simple expression language and extension.

@@

It is straightforward to model ASTs by inheritance and write corresponding parsers for all cases. Note that we have imported
the Scala Packrat Parsing library in code below, and use \lstinline{Parser} as a type synonym for \lstinline{PackratParser} throughtout the paper. \huang{This is the first time to demonstrate code in this paper. Readers may not know what those combinators/types are. Refer to the library docs here? move the 'scala parsing library' section to the beginning as 'background'?}

@@

\haoyuan{trait Expr with pretty printing? Shall we add client code here?}

\noindent As the combination of \inlinecode{pVar}, \inlinecode{pApp} and \inlinecode{pParen}, \inlinecode{pExpr} can parse expressions like \lstinline{"x (y z)"} correctly.

\paragraph{Attempt to Extend the Parser} Now consider we are extending the abstract syntax as well as the parsers. Hence we introduce lambda abstractions as a new case, which refers to the famous \textit{untyped lambda calculus}.

@@

The corresponding AST is extended as below.

@@

\lstinputlisting[linerange=26-31]{../Scala/Parser/src/PaperCode/Overview/Overview.scala}% APPLY:linerange=OVERVIEW_SIMPLE_EXT

@@

\paragraph{Problem: Hard-coded Recursive Calls} The critical problem of extending parsers directly is that the recursive calls in old parsers are hard-coded. Specifically, \inlinecode{pApp} makes two recursive calls to parse two consecutive expressions as an application. Before the extension, we implemented it by using \inlinecode{pExpr} for those recursive calls. It worked fine at that time because \inlinecode{pExpr} covers all cases of the old grammar. However, after the extension of lambda abstractions, the proper function for such recursive calls should be \inlinecode{pExtExpr} instead of \inlinecode{pExpr}, because the new function \inlinecode{pExtExpr} includes the lambda case in the new grammar. \lstinline{pParen} has the similar problem.

That is the reason why the extended parser fails to parse some valid expressions such as \inlinecode{(\\x.x)} \inlinecode{(\\y.y)}. To resolve this issue, \inlinecode{pApp} and \inlinecode{pParen} need to be aware of the change of grammar. We have some direct solutions:

\begin{itemize}
    \item Keep the old code, rewrite non-extensible functions like \inlinecode{pApp} and \inlinecode{pParen}. \\Separate compilation is obtained but code reuse is lost, because every function may contain such recursive calls.
    \item Modify the old code, correct the recursive calls by proper functions. \\We can avoid code duplication so as to get code reuse, but separate compilation is sacrificed because of modification on the old code.
    \item ??? \huang{third one? use dynamic languages?}
\end{itemize}

Because of the trade-off, neither of them are satisfying. The challenge of building semantically modular parsers requires fundamental change of the parser structure. We will introduce our solution in the next sections.

@@

Hard-coded recursive calls prevent our parsers from extensions.
This problem can be solved using delegation, which can be encoded with
\textit{open recursion}. The idea is simple: instead of making direct
recursive calls, an additional argument, that abstracts over the
recursive call, is passed to the parsers. This technique enables the
development of parsers that support OO-style extensibility of ASTs.

@@

\lstinputlisting[linerange=52-62]{../Scala/Parser/src/PaperCode/Overview/Overview.scala}% APPLY:linerange=OVERVIEW_OPEN_BASE

According to the definition of type synonym \inlinecode{Open[T]},
every parsing function takes a parser of \inlinecode{Expr} as a
parameter, namely \inlinecode{self}. This parameter \inlinecode{self}
represents \emph{the parser} of expressions, which is dynamiclly
explained by the argument. Using an abstract parameter provides
extensibility to parsers, because the recursive calls are not
hard-coded here --- they are dynamic and can be changed corresponding
to the grammar!

@@

\paragraph{Client Code} Since the type of parsing functions is
changed, we can not apply them directly. A \inlinecode{fix} function, which is the
the standard \textit{fixpoint combinator}, is used to `close' the open
recursion. After that, it can be used as an ordinary parser.
The client code below demonstrates the usage of new parsers.

@@

\lstinputlisting[linerange=67-71]{../Scala/Parser/src/PaperCode/Overview/Overview.scala}% APPLY:linerange=OVERVIEW_OPEN_USE

@@

\paragraph{Extensibility} Similarly, we extend the grammar by adding lambda abstractions as a new case. The new parser also extends the old one by adding \inlinecode{pLam} on it. Notice that the old code is not touched, hence separate compilation is obtained.

\lstinputlisting[linerange=76-82]{../Scala/Parser/src/PaperCode/Overview/Overview.scala}% APPLY:linerange=OVERVIEW_OPEN_EXT

\haoyuan{Code without combinators?}

@@

\lstinputlisting[linerange=88-88]{../Scala/Parser/src/PaperCode/Overview/Overview.scala}% APPLY:linerange=OVERVIEW_OPEN_EXT_USE

As we demonstrated, delegation encoded by open recursion is the key technique to obtain semantic modularity. It enables type-safe code reuse and separate compilation in our parsers. We will discuss this topic further in Section \ref{sec:openandparsing}.

### 2016-12-21 paper/sections/RelatedWork.tex
Update the tex template. Minor comments on Overview.

\paragraph*{Syntactically Extensible Parsing} Extensible parsing is achieved in many
different areas, of which parser generators are a mainstream area
specially designed for modular syntax and parsing. Many parser
generators~\cite{antlr1995,Grimm2006,Gouseti2014,Warth2016} support
modular grammars. They allow users to create new
modules where new non-terminals and production rules can be
introduced, some can even override existing rules in the old grammar
modules. For instance, \textit{Rats!}~\cite{Grimm2006} constructs its
own module system for the collection of grammars, while
NOA~\cite{Gouseti2014} uses Java annotation processing to gather
information together. Those parser combinators focus on the
\textit{syntactic extensibility} of grammars, and rely on a whole
compilation to generate a global parser, even with a slight
modification. Some of them may statically check the correctness and
unambiguity of grammars, but separate compilation and modular
type-checking remain unsolved, together with the issue of performance.
\bruno{what performance issue?}

Macro systems like C preprocessor, C++ templates and
Racket~\cite{Tobin-Hochstadt2011}, and other meta-programming
techniques are a similar area aiming at syntactic extensibility. 
SugarJ~\cite{Erdweg2011} conveniently introduces syntactic sugar
for Java using library imports. Composition of syntactic
sugar is easy for users, but it requires many rounds of parsing
and adaption. This significantly affects efficiency of compilation,
moreover, the implementation was based on SDF~\cite{Heering1989} and
Stratego~\cite{Visser2001}, which focused little on separate
compilation\bruno{focused little or did not have it at all?}. 
Extensible compilers like JastAdd~\cite{Ekman2007} and
Polyglot~\cite{Nystrom2003} also support extensible parsing, but this
is mostly done using parser generators. Moreover 
they focus on the extensions to a host language. Our library
is designed for modular language parsers in a type-safe way, with
flexible language composition. Although overriding existing production
rules is tricky with our approach, we could perhaps make use of
embedded domain-specific languages on top of parser combinators and
transformations for overriding, but it is anyway an orthogonal issue.
\bruno{Not sure that I understand the last sentence}
\haoyuan{???} \haoyuan{BTW what about Racket?} \bruno{Have you read 
Languages as Libraries? That is probably an important reference, which
I think we should cite.}\haoyuan{what about
  metafront? it is a macro system but does it have type safety?}
\haoyuan{Extensible syntax with lexical scoping?} \haoyuan{"Extensible
  syntax" proposes a system for extensible syntax, where users write
  EDSLs in their language with concrete syntax. Users can write rules
  for type-based disambiguation. But separate compilation is again not
  mentioned. Shall we mention that thesis?} \haoyuan{attribute
  grammars?}

\paragraph{Extensible Parsing Algorithms}
Extensible parsing algorithms are another area that indeed relates to
separate compilation. Specifically, \cite{Bravenboer2009} introduces
\textit{parse table composition}, in which paper grammars\bruno{what
  is a paper grammar?} are compiled
to the generation of parse tables\bruno{I don't understand the last
  part; please rephrase. Also please break the sentence into multple
  sentences. Generally speaking go over the text you wrote and watch
  out for long sentences. You have a tendency to write very long
  sentences. }, which are DFAs or NFAs, later they
can be composed by an algorithm, so as to provide separate compilation
for parsing. Nevertheless, the generation of parse tables is rather
expensive\bruno{expensive in terms of what? why is it expensive?}. 
Furthermore, \name supports \emph{both} separate compilation
as well as \emph{modular type-checking}. The extensibility of
parsing in our approach is further available at language composition
and lexical level. \haoyuan{I mentioned one shortcoming of our
  approach here, which is we cannot override existing production
  rules; another one is that we do not have explicit
  correspondence/relationship between abstract syntax and the parser.}

\paragraph*{Parser Combinators} Since~\cite{burge1975} firstly
introduced parser generators, and after Wadler discussed more details
on backtracking in~\cite{Wadler1985}, parser combinators have been
more and more popularly developed and used in the research area of
parsing. Among them many parsing libraries produces recursive descent
parsers by introducing functional, or more specifically, monadic
parser combinators with a foundation on~\cite{nott237}. Related work
includes Parsec~\cite{Leijen2001}, which is frequently used in Haskell
for context-sensitive grammars with infinite lookahead. Nevertheless,
left-recursion is known as a big issue in recursive descent parsers,
in which case programmers using libraries like Parsec have to manually
eliminate left-recursion in the grammar, which is cumbersome, but also
it distorts the shape of its old grammar, restricting the modularity
of parsing to extreme extent. It is possible to avoid some of  
 the limitations of parser combinators. 
\bruno{I'm not sure that th following text should be here. It could 
be in the paper, but somewhere else.}
For example, we have
conducted our previous experiment in Haskell, by using open recursion,
MRM~\cite{Oliveira2015} and Parsec. To support direct left-recursive
grammars, we made use of the state monad in Parsec to ensure that a
left-recursive production rule will not be applied successively in
parsing. It turned out to be successful, yet introduced complexity in
programming, hence we believe that the parsing library is obligated
for left-recursion.

@@

\paragraph*{Extensibility} There has been alot of work on building
modular programs, in both functional and object-oriented programming .
For instance, the famous DTC paper~\cite{swierstra2008} represents
modular ASTs using co-products of every two functors, which forms a
binary tree structure. It supports extensible datatypes, as well as
the operations on them by implementing instances of functors. Later
Oliveira et al. presented MRM~\cite{Oliveira2015} in Haskell, which is
similar to DTC but adopts a list of functors in type-level lists
instead. As mentioned above, we have conducted our early experiment
with MRM for modular datatypes, whereas the complexity of library
implementation was comparably high. Other related work
includes~\cite{Bahr2014}, where data types are flexible with
composition as well as decomposition.

>>>>

\paragraph{Syntactically Extensible Parsing} Extensible parsing is achieved in many
different areas, of which parser generators are a mainstream area
specially designed for modular syntax and parsing. Many parser
generators~\cite{antlr1995,Grimm2006,Gouseti2014,Warth2016} support
modular grammars. They allow users to create new
modules where new non-terminals and production rules can be
introduced, some can even override existing rules in the old grammar
modules. For instance, \textit{Rats!}~\cite{Grimm2006} constructs its
own module system for the collection of grammars, while
NOA~\cite{Gouseti2014} uses Java annotation processing to gather
information together. Those parser combinators focus on the
\textit{syntactic extensibility} of grammars, and rely on a whole
compilation to generate a global parser, even with a slight
modification. Some of them may statically check the correctness and
unambiguity of grammars, but separate compilation and modular
type-checking remain unsolved, together with their affects on the performance.
\bruno{what performance issue?}\haoyuan{Revised}

Macro systems like C preprocessor, C++ templates and
Racket~\cite{Tobin-Hochstadt2011}, and other meta-programming
techniques are a similar area aiming at syntactic extensibility.
SugarJ~\cite{Erdweg2011} conveniently introduces syntactic sugar
for Java using library imports. Composition of syntactic
sugar is easy for users, but it requires many rounds of parsing
and adaption. This significantly affects efficiency of compilation,
moreover, the implementation was based on SDF~\cite{Heering1989} and
Stratego~\cite{Visser2001}, which focused little on separate
compilation\bruno{focused little or did not have it at all?}.\haoyuan{The authors have
mentioned separate compilation as a possible yet unimportant feature; they decided to discard it.}
Extensible compilers like JastAdd~\cite{Ekman2007} and
Polyglot~\cite{Nystrom2003} also support extensible parsing, but this
is mostly done using parser generators. Moreover
they focus on the extensions to a host language. Our library
is designed for modular language parsers in a type-safe way, with
flexible language composition. Although overriding existing production
rules is tricky with our approach, we could perhaps make use of
embedded domain-specific languages on top of parser combinators and
transformations for overriding, but it is anyway an orthogonal issue.
\bruno{Not sure that I understand the last sentence}
\haoyuan{???} \haoyuan{BTW what about Racket?} \bruno{Have you read
Languages as Libraries? That is probably an important reference, which
I think we should cite.}\haoyuan{what about
  metafront? it is a macro system but does it have type safety?}
\haoyuan{Extensible syntax with lexical scoping?} \haoyuan{"Extensible
  syntax" proposes a system for extensible syntax, where users write
  EDSLs in their language with concrete syntax. Users can write rules
  for type-based disambiguation. But separate compilation is again not
  mentioned. Shall we mention that thesis?} \haoyuan{attribute
  grammars?}

\paragraph{Extensible Parsing Algorithms}
Extensible parsing algorithms are another area that indeed relates to
separate compilation. Specifically, \cite{Bravenboer2009} introduces
\textit{parse table composition}, where grammars\bruno{what
  is a paper grammar?}\haoyuan{Revised} are compiled
to the generation of parse tables\bruno{I don't understand the last
  part; please rephrase. Also please break the sentence into multple
  sentences. Generally speaking go over the text you wrote and watch
  out for long sentences. You have a tendency to write very long
  sentences. }, which are DFAs or NFAs, later they
can be composed by an algorithm, so as to provide separate compilation
for parsing. Nevertheless, the generation of parse tables is rather
expensive\bruno{expensive in terms of what? why is it expensive?}.
\haoyuan{in terms of performance. The sentence is copied from original paper.}
Furthermore, \name supports \emph{both} separate compilation
as well as \emph{modular type-checking}. The extensibility of
parsing in our approach is further available at language composition
and lexical level. \haoyuan{I mentioned one shortcoming of our
  approach here, which is we cannot override existing production
  rules; another one is that we do not have explicit
  correspondence/relationship between abstract syntax and the parser.}

\paragraph{Parser Combinators} Since~\cite{burge1975} firstly
introduced parser generators, and after Wadler discussed more details
on backtracking in~\cite{Wadler1985}, parser combinators have been
more and more popularly developed and used in the research area of
parsing. Among them many parsing libraries produces recursive descent
parsers by introducing functional, or more specifically, monadic
parser combinators with a foundation on~\cite{nott237}. Related work
includes Parsec~\cite{Leijen2001}, which is frequently used in Haskell
for context-sensitive grammars with infinite lookahead. Nevertheless,
left-recursion is known as a big issue in recursive descent parsers,
in which case programmers using libraries like Parsec have to manually
eliminate left-recursion in the grammar, which is cumbersome, but also
it distorts the shape of its old grammar, restricting the modularity
of parsing to extreme extent. It is possible to avoid some of
 the limitations of parser combinators.
\bruno{I'm not sure that th following text should be here. It could
be in the paper, but somewhere else.}
For example, we have
conducted our previous experiment in Haskell, by using open recursion,
MRM~\cite{Oliveira2015} and Parsec. To support direct left-recursive
grammars, we made use of the state monad in Parsec to ensure that a
left-recursive production rule will not be applied successively in
parsing. It turned out to be successful, yet introduced complexity in
programming, hence we believe that the parsing library is obligated
for left-recursion.

@@

\paragraph{Extensibility} There has been alot of work on building
modular programs, in both functional and object-oriented programming .
For instance, the famous DTC paper~\cite{swierstra2008} represents
modular ASTs using co-products of every two functors, which forms a
binary tree structure. It supports extensible datatypes, as well as
the operations on them by implementing instances of functors. Later
Oliveira et al. presented MRM~\cite{Oliveira2015} in Haskell, which is
similar to DTC but adopts a list of functors in type-level lists
instead. As mentioned above, we have conducted our early experiment
with MRM for modular datatypes, whereas the complexity of library
implementation was comparably high. Other related work
includes~\cite{Bahr2014}, where data types are flexible with
composition as well as decomposition.
